---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# terminaldigits

<!-- badges: start -->
<!-- badges: end -->

The package `terminaldigits` implements simulated tests of uniformity and independence for terminal digits. For certain parameters, `terminaldigits` also implements Monte Carlo simulations for type I errors and power for the test of independence. Simulations are run in C++ utilizing Rcpp. 

## Installation

You can install the development version of `terminaldigits` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("josh-mc/terminaldigits")
```

## Usage

In many cases, terminal digits can be assumed to be uniformly distributed and independent of preceding digits. A violation of either of these assumptions may point to a data quality issue. 

The following examples are based on a data set taken from the third round of a decoy experiment involving hand-washing purportedly carried out in a number of factories in China. For details, see `decoy` and Yu, Nelson, and Simonsohn (2018).

The `td_uniformity` function tests the assumption of uniformity using Pearson's chi-squared statistic for goodness-of-fit.

```{r example}
library(terminaldigits)

td_uniformity(decoy$weight, decimals = 2, reps = 1000)
```

The `td_independence` function tests the assumption of independence. The default statistic is again Pearson's chi-squared statistic but the log-likelihood ratio statistic, the Freeman-Tukey statistic, and the root-mean-square statistic are also available. 

```{r}

td_independence(decoy$weight, decimals = 2, reps = 1000)
```

The `td_test` function is a wrapper for the above two functions. For more details, including a discussion of the `td_simulate` function, see the package introduction vignette. 

## References 

Yu, F., Nelson, L., & Simonsohn, U. (2018, December 5). “In Press at Psychological Science: A New 'Nudge' Supported by Implausible Data.” DataColoda 74. http://datacolada.org/74
