---
title: "Package Introduction"
output: rmarkdown::html_vignette
author: Josh McCormick
date: May 2, 2022
vignette: >
  %\VignetteIndexEntry{packages_introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(terminaldigits)
```

The package `terminaldigits` implements simulated tests of uniformity and independence for terminal digits. `terminaldigits` also implements Monte Carlo simulations for type I errors and power (under certian deviations) for the test of independence. Simulations are run in C++ utilizing Rcpp. 

## Uniformity

When numbers are recorded with sufficient precision, for a wide range of data generating processes, terminal digits are uniformly distributed. Deviations from this distribution could indicate data quality issues. `terminaldigits` utilizes Pearson's chi-squared test of goodness-of-fit to assess the hypothesis that terminal digits are uniformly distributed. Rather than rely on the asymptotic approximation to the chi-squared distribution, `terminaldigits` utilizes the simulated chi-squared GOF test from the package `discretefit`.









